\documentclass[12pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{vmargin}
\usepackage{pslatex}
\usepackage{psfig}
%\setpapersize{USletter}
%\setmargins{1in}{1in}{6.3in}{8.8in}{0in}{0in}{0in}{0in}

\setlength{\footskip}{0.2in}
\setlength{\headsep}{0.1in}
\setlength{\oddsidemargin}{.6in}
\setlength{\topmargin}{1in}
\setlength{\textwidth}{7.1in}
\setlength{\textheight}{9.5in}
%\setlength{\headsep}{}

\renewcommand{\baselinestretch}{0.9}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%% begin document %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}


\section{Introduction}

This is a short step-by-step description of how to process the data in
bert-functional using the FreeSurfer Functional Analysis STream
(FS-FAST). Anatomical and functional data have been collected from a
subject code-named "bert". This is not meant to be an exhaustive
tutorial on all the functionality in FS-FAST but rather a means to get
users started on how to use FS-FAST, including interfacing with
FreeSurfer features such as registration, rendering functional data in
the volume using tkmedit as well as on the cortical surface using
tksurfer. For more information on using FS-FAST, consult
\$FREESURFER\_HOME/fsfast/docs, especially
MGH-NMR-StdProc-Handbook.doc.\\

\section{Unpacking}

This data can be found in the FreeSurfer distribution in a file called
bert.func.tar.gz. To begin processing, cd to the directory where you
want the functional data to be stored. This directory will be called
the Session Parent. Run the following command from the session parent
directory:
\begin{verbatim}
   tar xvfz <fsdist>/bert.func.tar.gz
\end{verbatim}
This will create a directory called bert-functional. This will have 4
subdirectories: \\

\begin{itemize}
\item {\bf 3danat} - 3d anatomicals collected with this session. These
are needed to automatically register the functionals with the
reconstructed anatomicals found in \$SUBJECTS\_DIR/bert

\item {\bf bold} -location of the functional runs. This will have three
subdirectories: 007, 008, and 009. Each of these will have a single
volume with stem "f" stored in bshort format. bold will also have a
file called seq.info in which information about the acquisition
sequence is held.

\item {\bf t1epi} - location of a t1-weighted EPI volume whose slice
prescription is identical to that of the functionals. The actual
volume is stored in a subdirectory called 004 in bshort format with a
stem called "f".

\item {\bf parfiles} - this is where the paradigm files for each run
are stored. The paradigm file gives information about which stimulus
was presented when (ie, the stimulus schedule). These files were
created by an FS-FAST program called optseq.
\end{itemize}

\section{Experimental Design}

The experiment involves an associative memory task using an
event-related design. The paradigm is described in detail in
\cite{Wag01}. This tutorial data was not part of that study, and that
study was not processed with FS-FAST. Russ Poldrack generously ran his
experiment using a subject who had agreed to provide data for this
tutorial.

The subject is shown one of 4 stimulus types. Each presentation lasted
3 seconds. Each stimulus has 5 words in the following pattern:

\begin{verbatim}
                         targetword
                   testword1     testword3
                   testword2     testword4
\end{verbatim}

The subject is asked to choose which test word is most related to the
target word.  By design, the target word is either highly related or
loosely related to one of the test words.  Sometimes there are four
test words and sometimes there are two test words and two "words" with
just X's in them.

This makes 4 conditions:
\begin{enumerate}
\item H2 - highly related, two words/two X's
\item H4 - highly related, four words
\item L2 - loosely related, two words/two X's
\item L4 - loosely related, four words
\end{enumerate}

This number coding is used in the paradigm files to identify which
stimulus was presented when.  The contrast (L2+L4) - (H2-H4) should
show activity in the left frontal region.

\section{Processing Steps}

\subsection{Create a Study Directory}

 The study directory is the "home base" for running processing
commands. The location of the study directory can be completely
independent from the directories where the functional data are
stored. It's best if you have only one study directory for a given
project regarless of how many sessions belong to the project. It also
avoids confusion if the study directory is NOT one of the session
directories.  All the programs are run from the study directory
regardless of which session is being processed.

\subsection{Create Session Id and Session Parent Files}

The Session Parent file is also known as the Session Directory file.
The Session Parent is the directory underwhich one or more sessions
are found. The root of the session is called the Session Id. In this
case, the Session Parent is the directory from which you run the tar
command to unpack the data; this is the full path (ie, starting with
/). The Session Id is "bert-functional".  Create an ascii file called
"sesspar" (the name can actually be anything) and set its contents to
the Session Parent. Create another file called "sessid" (its name can
also be anything) and set its contents to the Session Id. The
processing programs will search all the directories listed in the
Session Parent file for all the sessions listed in the Session Id
file.

\subsection{Copy the paradigm files into the functional run
directories}

cd into bert-functional.  There will be three paradigm files in a
subdirectory called "parfiles": sem\_assoc-s001-r001.dat,
sem\_assoc-s001-r002.dat, sem\_assoc-s001-r003.dat, one for each
run. The paradigm file reprensents which stimulus was presented when
during the run. The paradigm file for a run must be copied into the
directory with the data for that run.  The paradigm file can be called
anything, but the name must be the same across runs. Here, we will
call it sem\_assoc.par:

\begin{verbatim}
  cp parfiles/sem_assoc-s001-r001.dat bold/007/sem_assoc.par
  cp parfiles/sem_assoc-s001-r002.dat bold/008/sem_assoc.par
  cp parfiles/sem_assoc-s001-r003.dat bold/009/sem_assoc.par
\end{verbatim}

\subsection{Motion Correction}

cd to the Study Directory and type:

\begin{verbatim}
     mc-sess -sf sessid -df sesspar
\end{verbatim}

This uses AFNI to motion correct the functional data to the first
volume of the first run. After completion, there will be a new volume
called fmc in each run directory. There will also be an ascii file
called fmc.mcdat with the motion correction parameters for each time
point. There are 10 columns in this file corresponding to: (1) time
point number, (2) rotation (deg), (3) rotation (deg), (4) rotation
(deg), (5) translation (mm), (6) translation (mm), (7) translation
(mm), (8) RMS error before correction, (9) RMS error after correction,
and (10) total (vector) translation (mm). Those who are using this
program should cite \cite{Cox99}.

\subsection{Intensity Normalization}

This stage is slightly misnamed because intensity normalization does
not actually occur here. Rather, it segments the tissue from the air
and computes the global mean of the fMRI signal inside the
tissue. This number is later used to rescale the data so that, when
intersubject averaging is done, all subjects have the same global
mean. To intensity normalize the motion corrected data, cd to the
Study Directory and run:

\begin{verbatim}
   inorm-sess -sf sessid -df sesspar -motioncor
\end{verbatim}

After processing, you will find a file called fmc.meanval in each run
directory. This is an ascii file in which the global mean is
stored. You will also find a file called fmc.report in which many
statistics relating to the "cleanliness" of your data are stored.

\subsection{Create Analysis}
\label{mkanal.sec}

An "analysis" is a collection of parameters used to analyze each
session. It encompasses the signal model, the stimulus schedule, and
any preprocessing options such as smoothing and motion correction. The
key here is to realize that you give a name to all these parameters
and then refer to that name in later processing stages. Creating the
analysis does not actually cause the data to be analyzed. The analysis
is created ONCE and then applied to any and all sessions. If you want
to change your analysis parameters, you should either create a new
analysis (with a new name) or delete the original and analysis and
create another one from scratch with the same name. For this tutorial,
create an analysis called "sem\_assoc" using the following command:

\begin{verbatim}
  mkanalysis-sess.new \
    -analysis sem_assoc-3b \
    -TR 2 \
    -paradigm sem_assoc.par \
    -designtype event-related \
    -funcstem fmc \
    -inorm \
    -fwhm 4 \
    -nconditions 4 \
    -timewindow 26 \
    -tprestim 4
\end{verbatim}

The name of the analysis is passed with the -analysis flag. The TR is
2 seconds. The name of the paradigm file is called sem\_assoc.par. This
is an event-related design as indicated by -designtype. The -funcstem
is the stem of the volume to be processed. In this case this is "fmc"
which is the volume created by the motion correction stage
above. -inorm instructs the processor to use the intensity
normalization file fmc.meanval created above to rescale the functional
data to a global mean of 1000. In-plane spatial smoothing is indicated
with the -fwhm flag (fwhm = full-width/half-max); in this case a fwhm
of 4 mm is used.  The number of conditions in the paradigm is given by
-nconditions; note that this does not include the null or fixation
condition (code 0 in the paradigm files).  The argument of the
-timewindow flag  has been set to 26 seconds indicating the time range
in an FIR signal model. The -tprestim 4 option instructs the processor
to begin the FIR time window 4 seconds before the stimulus onset. This
step will create a subdirectory in the Study Directory with the given
analysis name. In that directory will be two files: analysis.info and
analysis.cfg with the parameters indicated on the command line.

\subsection{Average the Data for the Session} 

Step \ref{mkanal.sec} created the analysis by collecting all the
parameters needed to analyse the data. The actual analysis is done by
the command:

\begin{verbatim}
   selxavg-sess -sf sessid -df sesspar -analysis sem_assoc
\end{verbatim}

This will separately analyze the data in each session indicated in the
sessid file. After completion, there will be a new subdirectory under
bold with the same name as the analysis. This directory will have two
volumes: h-offset and h. h-offset is a map of the mean value at each
voxel. h contains the estimated hemodynamic response at each voxel as
well as the standard devation of the residual error. There is also an
ascii file called h.dat with information about the parameters using
during the analysis.

\subsection{Create Omnibus Contrast}
\label{mkomnibus.sec}

A {\em contrast} is an instantiation of a hypothesis.  The omnibus
contrast is a test for any task-related activity against the
baseline. Here, the "baseline" means the variance left unexplained
after fitting the time course at each voxel for the mean, linear
trend, and task-related activity.  Mathematically, the mean of the
baseline is forced to zero. The omnibus is usually tested first to
assure that the data have been processed properly and that the subject
is responding. If there is no omnibus activity, then there is no use
in continuing the analysis. The omnibus is created using the command:

\begin{verbatim}
  mkcontrast-sess \
     -analysis sem_assoc \
     -contrast omnibus \
     -a 1 -a 2 -a 3 -a 4 -c 0 \
     -nosumconds
\end{verbatim}

As with mkanalysis-sess, making the contrasts is not the same as
computing the contrast at each voxel. Making the contrast just
collects all the information needed to compute the contrast. A
contrast only needs to be made once. If you change the analysis (ie,
the time window, the prestimulus window, or the number of conditions),
the contrast will need to be remade. After making the contrast, you
will see a file in the analysis directory called omnibus.mat.

\subsection{Compute the Omnibus Contrast}

The actual contrasts are computed with stxgrinder-sess:

\begin{verbatim}
  stxgrinder-sess -contrast omnibus \
      -analysis sem_assoc \
      -sf sessid -df sesspar 
\end{verbatim}

This will create a directory called
bert-functional/bold/sem\_assoc/omnibus in which several volumes will
be stored. Each volume corresponds to a different type of statistical
test and so represents a different statistical map. These volumes are:
f (value of the F ratio), fsig (significance of the F-test), t (value
of the t ratio), sig (significance of the t-test, possibly mulitple
maps for multiple post-simulus delays), minsig (the single best
signifiance of the t-test (bonferroni corrected) at each voxel), and
iminsig (the index of the best t-test signifiance at each
voxel). These will be elaborated below.

\subsection{View the Contrast on the Slices}
\label{sliceview.sec}

\begin{verbatim}
  sliceview-sess -contrast omnibus \
      -analysis sem_assoc \
      -sf sessid -df sesspar \
      -map fsig -slice mos -nohdr
\end{verbatim}

This displays all the slices as a mosaic. The base image is the
average functional image. The overlay is the signifiance of the
F-test. Color indicates signifiances in the range of 2 (ie, $10^-2$) to
7 (ie, $10^-7$). The red/yellow scale indicates activations above
baseline; the blue scale indicates activations above below baseline.
Other maps (ie, f, sig, minsig, iminsig) can be viewed by changing the
-map option (see below). For the omnibus test, only the fsig is
relevant. If one wants to view only a single slice, use -slice
SLICENO, where SLICENO is the zero-based slice number. The -nohdr
option tells sliceview not to load the hemodyanmic response (see
below).

\subsection{Viewing the Hemodynamic Responses}

The hemodynamic responses could have been viewed with the omnibus
contrast, but it is more instructive to view them using an
all-versus-baseline overlay. First, create the contrast:

\begin{verbatim}
  mkcontrast-sess \
     -analysis sem_assoc \
     -contrast allvbase \
     -a 1 -a 2 -a 3 -a 4 -c 0
\end{verbatim}

Note that this is similar to the omnibus created in Step
\ref{mkomnibus.sec} except that it does not include the -nosumconds
flag. This means that the contrast will sum the conditions together
whereas the omnibus tests each condition separately. The omnibus and
all-v-baseline generally paint the same picture, but all-v-baseline
allows for some more sophisticated display options. Compute the
contrast maps:
\begin{verbatim}
  stxgrinder-sess -contrast allvbase \
      -analysis sem_assoc \
      -sf sessid -df sesspar 
\end{verbatim}
View the contrast maps:
\begin{verbatim}
  sliceview-sess -contrast allvbase \
      -analysis sem_assoc \
      -sf sessid -df sesspar \
      -map sig -slice mos 
\end{verbatim}

Note that the -nohdr flag has been removed which forces the
hemodynamic averages to be loaded (this can take a while). Also the
map is now sig instead of fsig, meaning that you are about to view the
per-poststimulus-delay t-significance maps. When the slices are
displayed, you will not see much activation. This is good because you
are looking at the map of the activation 4 seconds *before* stimulus
onset. Click in the window and hit ``g''. This will bring up another
window with four plots; each plot represents the unbiased hemodynamic
response to each of the four conditions at the voxel selected in the
image window. As you click in the image window, you will change the
time courses in the plot window. Click in the plot window and hit
``e''; this will display the standard error bars for each condition at
each time point.  Click in the image window and hit ``+''. This will
advance the map from that of 4 seconds before stimulus onset to 2
seconds before stimulus onset.  You will also see a vertical line in
the plot window appear at -2 sec. This bar indicates which
post-stimlus delay you are currently viewing. Hitting ``+'' again will
advance to the next frame thus allowing you to view the activation
like a movie. Note, to get help with the keypress commands, click in
either window and hit ``h''.

\subsection{A Contrast-of-Interest}

Upto this point, you have only been looking at task-vs-nothing
contrasts, which, while useful, are not particularly interesting. This
contrast will compare the response to two tasks, loosely related with
4 words (L4) versus highly related with 2 words (H2):

\begin{verbatim}
  mkcontrast-sess \
     -analysis sem_assoc \
     -contrast L4vH2 \
     -a 4 -c 1 -sumdelays
\end{verbatim}

Using ``-a 4 -c 1'' means that positive values indicate that L4 is
greater than H2. Conditions H4 and L2 are ignored.  Normally, a
separate t-test is done at each poststimulus delay. However, the
-sumdelays flag indicates that the statistical test should be done
after summing the hemodynamic responses across poststimulus
delays. This can help bring out activation when the responses to two
conditions are similar but one is slightly and consistently larger
than the other over time. Compute and view the contrast:
\begin{verbatim}
  stxgrinder-sess -contrast L4vH2 \
      -analysis sem_assoc \
      -sf sessid -df sesspar 
  sliceview-sess -contrast L4vH2 \
      -analysis sem_assoc \
      -sf sessid -df sesspar \
      -map sig -slice mos 
\end{verbatim}
In the $9^{th}$ slice, you should see a small patch of activation in
the {\em posterior} portion of the Left Inferior Prefrontal Cortex
(LIPC).  In the $13^{th}$ slice, you should see a small patch of
activation in the {\em anterior} portion of the LIPC.  Note that the
images are in radiological convention, so the activity will appear on
the right side of the image. Also note that it is not possible to
scroll through the different post-stimulus delays because they have
all been collapsed to one image.

\section{Interfacing with FreeSurfer}

Linking into FreeSurfer allows the user to view the functional results
on high-resolution 3D anatomical images as well as on the cortical
surface. It also allows the user to resample into talairach,
spherical, or region-of-interest spaces in order to perform
intersubject averaging. Of course, to link to FreeSurfer, the subject
must have been processed through the FreeSurfer anatomical processing
stream (see \$FREESURFER\_HOME/docs/FreeSurferManual.pdf). 

\subsection{The Subjectname File}

One of the first steps in the anatomical processing stream is to
create a subject directory using the command {\em mksubjdirs
subjectname} where ``subjectname'' is a unique identifier for the
subject that will be analyzed. This is also known as the {\em Subject
Identifier String }. The subject's anatomical directory must be
visible from \$SUBJECTS\_DIR (ie, you should see something when
running the unix command {\em \$SUBJECTS\_DIR/subjectname}). To
interface FS-FAST to FreeSurfer, you must create a file called
subjectname in the functional session directory the contents of which
should be the name of the subject. Note that it is the contents that
hold the subject identifier string. The name of the file should
actually be ``subjectname''. For this tutorial, go to bert-functional
and create a file called subjectname with conents ``bert'' (no
quotes). Note, run {\em ls \$SUBJECTS\_DIR/bert} to assure that bert
exists on your system.

\subsection{Functional/Anatomical Registration}

The functional data and the anatomical data are collected using
different resolutions, fields-of-view, and contrast weightings. Often,
they are collected on different scanning sessions, sometimes years
apart, and on different scanners. Yet, we still need to assign voxels
in the anatomical space to voxels in the functional space in order to
view the functional data on the anatomical.  This is done in three
stages, two of which are automatic.

\subsubsection{Automatic Registation}

The automatic registration can be used only if a high-resolution
T1-weighted anatomical was acquired during the same scanning
session. These may have been collected in order to process the subject
with FreeSurfer, in which case there will probably be 2 or 3 very
high-quality scans. Or, if a subject has already been reconstructed,
there may only be one low-quality anatomical. The latter is the case
for bert. One run of low-quality anatomicals can be found in
bert-functional/3danat.  These are referred to as the ``same-session
anatomicals'' where as the those in \$SUBJECTS\_DIR/bert are referred
to as the ``database anatomicals''. As mentioned above, the automatic
registration is performed in two stages. First, the transform between
the same-session anatomicals and the functional is computed based on
the information about the field-of-view of each scan. Second, the
transform between the same-session anatomicals and the database
anatomicals is computed by performing a voxel-by-voxel comparison
between the two and adjusting the positions so that the two volumes
match best. Concatenating these two transforms yields the tranform
between the functionals and the database anatomicals.  Those using
this procedure should cite \cite{Collins94}.  This is accomplished
using the command:

\begin{verbatim}
  autoreg-sess  -sf sessid -df sesspar 
\end{verbatim}

This will create a file called register.dat in
bert-functional/bold. This file will have 8 lines. The first four
lines are: subjectname, in-plane resolution, between-plane resolution,
and intesity value. The between-plane resolution is the slice
thickness if there is no skip. The intensity value does not actually
have anything to do with the registration itself; it is used when
displaying the functional volume during the manual registration
phase. The last four lines of register.dat have the values of the
registration matrix that converts a location in anatomical space to a
location in functional space.

\subsubsection{Manual Registation}

Manual registration is used to check and fine-tune the automatic
registration or to perform the entire registation in the case that
there are no same-session anatomicals or if the automatic registration
fails. Even if the automatic registration succeeds, {\em always check
the registation manually} using the following command:
\begin{verbatim}
  tkregister-sess -sf sessid -df sesspar 
\end{verbatim}
This will bring up two windows, one with a brain, the other a control
window. Hitting the ``Compare'' button in the control window will
cause the image window to alternate between the functional and the
anatomical volumes. Change the view by hitting the ``Coronal'' or
``Sagital'' or ``Horizonal'' buttons. The functional brain can be
translated left or right by adjusting the horizontal scroll bar under
``Translate Brain''. It can be translated up and down using the
vertical scroll bar. The brain can be rotated about the red cross
using the ``Rotate Brain'' scroll bar. After adjusting the
registration, hit the ``SAVE REG'' button (answer yes to overwrite).

\subsubsection{Registration Used in Tutorial}

Here are the contents of register.dat used in this tutorial:
\begin{verbatim}
bert
3.124990
6.000000
0.200000
 1.003765e+00 -6.999285e-02  8.023400e-02  3.269092e+00 
 9.321700e-02  2.028884e-01 -9.838270e-01  1.762333e+01 
 5.287400e-02  9.710374e-01  2.113440e-01  1.254106e+01 
 0.000000e+00  0.000000e+00  0.000000e+00  1.000000e+00 
\end{verbatim}

\section{Viewing the Functional Results on the Anatomical}

In section \ref{sliceview.sec}, we showed how to view the functional
results overlayed on the original functional slices. In this section
we show how to view the functional results on the high-resolution T1
anatomical volume.

\begin{verbatim}
   tkmedit-sess -sf sessid -df sesspar -a sem_assoc -c omnibus -map fsig
\end{verbatim}

You should see activation in the posterior LIPC at approximately
talairach coordinates $(-54, 16, +28)$.  You should see activation in
the anterior LIPC at approximately talairach coordinates $(-58, 12,
+2)$. There should also be activation in visual and motor cortices.\\

\section{Viewing the Functional Results on the Surface}

\begin{verbatim}
  paint-sess -sf sessid -df sesspar -a sem_assoc -c omnibus -map fsig
  surf-sess -sf sessid -df sesspar -a sem_assoc -c omnibus  -map fsig
\end{verbatim}


%-------------------------------------------------------------------%
\begin{thebibliography}{99}

\bibitem[Collins, et al, 1994] {Collins94} 
D. L.  Collins, P.  Neelin, T.  M.  Peters and A.
C.  Evans, Automatic 3D Inter-Subject Registration of MR Volumetric
Data in Standardized Talairach Space, Journal of Computer Assisted
Tomography, 18(2) p192-205, 1994.

\bibitem[Cox, 1996]{Cox96}
RW Cox; AFNI: Software for analysis and visualization of functional
magnetic resonance neuroimages; Computers and Biomedical Research, 29:
162-173, 1996.

\bibitem[Cox and Jesmanowicz, 1996]{Cox99} RW Cox and A Jesmanowicz.
Real-time 3D image registration for functional MRI.  Magnetic
Resonance in Medicine, 42: 1014-1018, 1999.

\bibitem[Wagner, et al, 2001]{Wag01} Wagner, A.D., Pare-Blagoev, J.,
Clark, J. and Poldrack, R.A. Recovering Meaning: Left
Prefrontal Cortex Guides Controlled Semantic Retrieval. Neuron
31:329-338. August 2, 2001.

\end{thebibliography}



\end{document}